---
title: "Bonus Question"
author: "Hikmat Paudel"
date: "2026-04-09"
output:
  word_document: default
  md_document: default
  pdf_document: default
---

# 1. Recap Material
 
Give me notes on the recap material. These questions focus on material that students commonly struggle with (data wrangling, plotting, modeling, or debugging).

##Data Wrangling (dplyr)

Data wrangling is the process of preparing and transforming raw data into a useful format. The `dplyr` package provides several important functions:

- `select()` is used to choose specific columns  
- `filter()` extracts rows based on conditions  
- `mutate()` creates or modifies variables  
- `summarise()` calculates summary statistics  
- `group_by()` groups the data before summarizing  

It is important to always check column names and ensure variables have the correct data types before performing operations.

Example

```{r}
library(dplyr)

# Create example dataset
df <- data.frame(
  year = c(2025, 2025, 2024, 2025),
  treatment = c("A", "B", "A", "B"),
  yield = c(10, 15, 8, 20),
  rep1 = c(5,6,4,7),
  rep2 = c(7,5,6,8)
)

# Perform filtering, grouping, and summarizing
data_summary <- df %>%
  filter(year == 2025) %>%
  group_by(treatment) %>%
  summarise(mean_yield = mean(yield, na.rm = TRUE))

# Show results
data_summary
```

Common Mistakes
Using summarise() without grouping first
Incorrect or misspelled column names
Treating categorical variables as numeric

##Plotting (ggplot2)
The ggplot2 package is used to create visualizations. A basic plot structure looks like this:

```{r}
library(ggplot2)

# Scatter plot with real columns
ggplot(df, aes(x = year, y = yield)) +
  geom_point() +
  theme_classic()
```
Common layers include:
geom_point() for scatterplots
geom_line() for line graphs
geom_bar(stat = "identity") for bar charts
facet_wrap() for multiple panels
theme_classic() for a simple design

Example
```{r}
ggplot(df, aes(x = treatment, y = yield)) +
  geom_boxplot() +
  theme_classic()
```

Common Mistakes
Incorrect placement of variables inside/outside aes()
Forgetting to use + between layers
Using wrong variable names
Plotting unsummarized data incorrectly

##Linear Modeling
Linear models are used to study relationships between variables.

Examples
```{r}
model <- lm(yield ~ treatment, data = df)
summary(model)
```

Important Outputs
Estimate: shows the effect size
p-value: indicates statistical significance
R-squared: measures model fit
Residuals: show prediction errors

Prediction
```{r}
predict(model)
```

Common Mistakes
Mixing up predictor and response variables
Not converting categorical variables into factors
Ignoring missing values in the dataset

##Debugging
Debugging helps identify and fix errors in code. Useful strategies include:
Carefully reading error messages
Running code step-by-step
Checking object and variable names
Verifying parentheses and commas
Inspecting data using str() and names()

Example
```{r}
str(df)
names(df)
```

# 2. Extra Topic Exploration

Complete one of the extra optional materials provided, OR explore a topic of your choice, or find a topic online and explore the code and how it works.

I explored the pivot_longer() function from the tidyr package, which is used to reshape data from wide format into long format. This is especially useful when preparing data for visualization and analysis.

Example

```{r}
library(tidyr)

long_df <- df %>%
  pivot_longer(
    cols = starts_with("rep"),
    names_to = "replicate",
    values_to = "value")

long_df
```

Why It Is Useful
Makes plotting easier
Simplifies grouping and summarizing
Helps maintain a tidy data structure

Justification
I chose this topic because reshaping data is an important step in many analyses. I wanted to better understand how to organize datasets in a way that works efficiently with visualization and modeling tools. This will help me work with more complex datasets in the future.

# 3. Introduction to renv

Follow the coding demonstration in class. Turn in notes.

## renv Notes

**Purpose:**
- Creates reproducible R environments  
- Keeps track of package versions  
- Helps others run the same code without issues  

## Initialize Project

```{r}
install.packages("renv")
renv::init()
```
Install Packages
```{r}
install.packages("dplyr")
```
Save Environment
```{r}
renv::snapshot()
```
Restore Environment
```{r}
renv::restore()
```

Files Created
renv.lock → stores package version details
renv/ folder → contains the project library

Why It Is Important
Ensures results can be reproduced
Avoids conflicts between package versions
Makes it easier to share projects with others



