---
title: "Codding Challenge 6"
author: "Hikmat Paudel"
date: "2026-03-26"
output:
  pdf_document:
    toc: true
  md_document:
    variant: gfm
  html_document:
    df_print: paged
  word_document:
---

# Question 1 (Regarding Reproducibility)
The main purpose of creating your own functions and using iteration is to keep your analysis consistent, reproducible, and efficient. Functions let you reuse code instead of rewriting it, which helps minimize errors and ensures the same approach is applied each time. Iteration tools like loops automate repetitive tasks, making it easier to apply your analysis to new or larger datasets.

# Question 2 (Conceptual)
1) Function in R:
A function in R is created by using the function() and assigning a name. Inputs are dfined in parentheses, and the code is written inside {}. It givess a result using return(). Functions are written in the scripts or R Markdown to resue code.

2) For loop in R:
A for loop repeats code multiple times using for (i in ...), where i goes through a sequence. The code inside{} runs for each value of i. Results are usually stored in a variable that updates each iteration.

# Question 3 (Load Data)
```{r}
Cities <-read.csv("Cities.csv")
```

# Question 4 (Function Writing)
```{r}
 Haversine<- function(lat1, lon1, lat2, lon2) {rad.lat1 <- lat1 * pi/180
  rad.lon1 <- lon1 * pi/180
  rad.lat2 <- lat2 * pi/180
  rad.lon2 <- lon2 * pi/180
  
  # Haversine formula
  delta_lat <- rad.lat2 - rad.lat1
  delta_lon <- rad.lon2 - rad.lon1
  a <- sin(delta_lat / 2)^2 + cos(rad.lat1) * cos(rad.lat2) * sin(delta_lon / 2)^2
  c <- 2 * asin(sqrt(a))
  
  # Earth's radius in kilometers
  earth_radius <- 6378137
  
  # Calculate the distance
  distance_km <- (earth_radius * c) / 1000
  
  return(distance_km) 
}
colnames(Cities)
```

# Question 5
```{r}
auburn <- Cities$lat[Cities$city == "Auburn" ]
auburn2 <- Cities$long[Cities$city == "Auburn" ]

nyc <- Cities$lat[Cities$city == "New York"]
nyc2 <- Cities$long[Cities$city == "New York"]

# Calculate distance 
 auburn_nyc_dist <- Haversine(auburn, auburn2, nyc, nyc2)

auburn_nyc_dist
```

# Question 6 (Loop)
```{r}
for (i in 1:nrow(Cities)) {
  
  dist <- Haversine(auburn, auburn2,
                    Cities$lat[i],
                    Cities$long[i])
  
  print(dist)
}
```

# Qusetion 7 (dataframe)
```{r}
distance_df <- NULL

for (i in 1:nrow(Cities)) {
    result_i <- data.frame(
      City1 = Cities$city[i],
      City2 = "Auburn",
      Distance_km = Haversine(auburn, auburn2,
                              Cities$lat[i],
                              Cities$long[i]))
    
    distance_df <- rbind.data.frame(distance_df, result_i)
}
print(distance_df)
```
# Question 8 (Github)
[CodingChallenge6]https://github.com/Hikmat-au/Assignments-repository 
