---
title: "Coding Challenge 5_Data Wrangling"
author: "Hikmat Paudel"
date: "2026-03-05"
output:
  pdf_document:
    toc: true
    number_sections: true
  md_document:
    variant: gfm
---
# Question 1 - Downloading and reading two csv files

```{r}
getwd()
library(tidyverse)
diversity<- read.csv("DiversityData.csv")
str(diversity)
metadata<- read.csv("Metadata.csv")
str(metadata)
```

# Question 2 - Joining the two dataframes together and creating a new dataframe named alpha

```{r}
library(dplyr)
alpha<-left_join(diversity, metadata, by="Code")
alpha
```

# Question 3 - Calculate Pielou’s evenness index

```{r}
alpha_even <- alpha %>%
  mutate(even = shannon / log(richness))
alpha_even
```

# Question 4 - Using tidyverse language of functions and the pipe
```{r}
library(tidyverse)
alpha_average <- alpha_even %>%
  group_by(Crop, Time_Point) %>%
  summarise(mean.even = mean(even, na.rm = TRUE),count = n(),sd.even = sd(even, na.rm = TRUE),se.even = sd.even / sqrt(count))
alpha_average
```

# Question 5 -  Calculating difference between soybean column,soil column, and between cotton column and soil column
```{r}
alpha_average2 <- alpha_average %>%
  select(Time_Point, Crop, mean.even) %>%
  pivot_wider(names_from = Crop,values_from = mean.even) %>%
  mutate(diff.cotton.even = Soil - Cotton,diff.soybean.even = Soil - Soybean)
alpha_average2
```

# Question 6 - Connecting it to plots
```{r}
library(tidyr)
library(ggplot2)

alpha_plot_long <- alpha_average2 %>%
  select(Time_Point, diff.cotton.even, diff.soybean.even) %>%
  pivot_longer(cols = c(diff.cotton.even, diff.soybean.even),names_to = "diff",values_to = "values")

plot<-ggplot(alpha_plot_long, aes(x = Time_Point, y = values, color = diff)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(x = "Time Point",y = "Evenness Difference",color = "Difference Type") +
  theme_minimal()
print(plot)
```

# Question 7 - My github assignment repository
https://github.com/Hikmat-au/Assignments-repository
